%!TEX root = ./hw3.tex
\qns{A Portfolio Design Problem}\rm
\label{exer:portfolio_socp}
The returns on $n = 4$ assets are described by a Gaussian (normal) random vector $r \in \mathbb{R}^n$ , having the following expected value $\hat{r}$ and covariance matrix $\Sigma$: 
\[
\hat{r} = 
\begin{bmatrix}
    0.12  \\
    0.10  \\
    0.07 \\
    0.03 
    
\end{bmatrix}, \; \Sigma = 
\begin{bmatrix}
    0.0064 & 0.0008 & -0.0011  & 0 \\
    0.0008 & 0.0025 & 0  & 0 \\
    -0.0011 & 0 & 0.0004  & 0 \\
    0 & 0 & 0  & 0
    
\end{bmatrix}
.
\]


The last (fourth) asset corresponds to a risk-free investment. An investor
wants to design a portfolio mix with weights $x \in \mathbb{R}^n$ (each weight $x_i$
is non-negative, and the sum of the weights is one) so as to obtain the best possible expected return $\hat{r}^Tx$, while guaranteeing that: \\ 

(i) No single asset weighs more than $40\%$. \\ 
(ii) The risk-free assets should not weigh more than $20\%$. \\ 
(iii) No asset should weigh less than $5\%$. \\ 
(iv) The probability of experiencing a return lower than $q = -1\%$ should be no larger than $\epsilon = 10^{-4}$.
\begin{enumerate}

\item {[Short justification, 1-2 lines]} For a scalar standard normal $y \sim \mathcal{N}(0, 1)$, show that the constraint

\begin{equation*}
    \textbf{Pr}(y \leq q) \leq \epsilon
\end{equation*}

can be written as

\begin{equation*}
    q \leq \Phi^{-1}(\epsilon),
\end{equation*}

where $\Phi(y)$ is the CDF of $y$, and $\Phi^{-1}$ is the inverse CDF. In particular, justify taking any inverse across inequality.

\newline
\sol{\input{portfolio_design_sol/1.tex}}
\newline

\item {[Short justification, 1-2 lines]} Given a multivariable Gaussian $r \sim \mathcal{N}(\hat{r}, \Sigma),r \in \mathcal{R}^n$, its projection onto any $x \in \mathcal{R}^n$ is also a scalar Gaussian. That is,

\begin{equation*}
    r^Tx \sim \mathcal{N}(\hat{r}^Tx, x^T \Sigma x).
\end{equation*}

Show that the fourth constraint, also known as the chance constraint, can be written as 

\begin{equation*}
\Phi^{-1}(10^{-4})||\Sigma^{1/2}x||_2 \geq -\hat{r}^Tx - 0.01.
\end{equation*}
Note that $\Phi^{-1}(10^{-4})$ is negative so this constraint can be converted to one in standard SOCP format as 
\begin{equation*}
||\Sigma^{1/2}x||_2 \leq \frac{1}{\Phi^{-1}(10^{-4})}(-\hat{r}^Tx - 0.01).
\end{equation*}

\emph{Hint: Write $r\tran x$ as a standard Normal.}

\newline
\sol{\input{portfolio_design_sol/2.tex}}
\newline

\item {[Optimization Problem Formulation]} Formulate the portfolio optimization problem as a SOCP.
Solve the problem using CVXPY in the Jupyter notebook. What is the maximal achievable expected return, under the above constraints? \\


% \textbf{Hint}: Constraint (iv) is known as a "chance constraint." 
% \[
% a^Tx \sim N(\hat{a}, \Sigma) \implies a^Tx - b \sim N(\hat{a}x - b, x^Tx\Sigma x).
% \]
% We then have:
% \[\Pr(a^Tx \leq b) \geq \eta \iff b - \hat{a}^Tx \geq \Phi^{-1}(\eta)||\Sigma^{1/2}x||_2
% \]

\newline
\sol{\input{portfolio_design_sol/3.tex}}
\newline

\item Solve the problem for a large number of values of $\epsilon$ between $10^{-6}$ and $10^{-1}$ and plot the optimal values of $\hat{r}^Tx$ versus $\epsilon$.   Also make an area plot of the optimal portfolios $x$ versus $\epsilon$. What do you observe as the risk tolerance $\epsilon$ decreases? Read  \href{https://en.wikipedia.org/wiki/Area_chart}{\color{blue} here} for what an area plot is. 

\newline
\sol{\input{portfolio_design_sol/4.tex}}
\newline

\item \textit{Monte Carlo simulation.} Let $x$ be the optimal portfolio found in part 1, with $\epsilon = 10^{-4}$ . This portfolio maximizes the expected return, subject to the probability of a loss being no more than 1 $\%$. Generate 10000 samples of $r$, and plot a histogram of the returns. Find the empirical mean of the return, and calculate the percentage of samples for which a loss occurs. For more on Monte Carlo, see \href{https://en.wikipedia.org/wiki/Monte_Carlo_method}{\color{blue} here}.

\newline
\sol{\input{portfolio_design_sol/5.tex}}
\newline

\end{enumerate}

